Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies

### 2018-09-19 legacy.tex
changed 'GWAS' and 'Q'

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/Figure1.jpg}
\caption{
\textbf{A} 
Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
The majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern. There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. These three enrichments can be summarized as G**${\rightarrow}$GG*. (note: the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
The histogram on the left of the plot is the distribution of significant variants. 
\textbf{C} 
Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1kGP. This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
The same analysis was performed independently for each of the populations in the 1kGP. }
 \label{SFS}
\end{figure}

			\section{Results}
%The results section is organized as follows.
First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
Finally, we investigate how these SNPs have impacted modern genomics analyses. 
We report that major imputation servers impute the dubious SNPs. 
We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.

@@

When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
Individuals with low mapping quality show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome wide simple logistic regression analysis (GWLR) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
In this case, we are looking for loci where the genotypes are predictable based on the data quality scores.
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.

Thus, this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques. 
The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the GWLR performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.

@@

\caption{\textbf{A}The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project. Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once. The x-axis is ranked by the mean sequencing date per population. \textbf{B} The x-axis is sorted by the sequencing date per individual. The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. The bottom plot indicates the sequencing technologies used over time.}
\label{MapQual}
\end{figure}

The GWAS approach similarly identified mapping-quality-associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}). We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.

@@

\caption{Overlap of SNPs identified independently to be associated with quality. 
The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
The line plot underneath shows the number of populations for which a variant has reached significance.
The populations that tend to have the most low-quality individuals also tend to have the most variants associated to quality. 
The same variants identified as being low quality independently in each population are found in other populations. }
  \label{OverLap}
\end{figure*}

\subsection{Suspicious SNPs impact modern genomics analyses}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
We found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. 
This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.

Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait. 
The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.

@@

\section{Discussion}
One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
We found that roughly 1\% of the SNPs associated to quality were found in HapMap; rs6057648 is among those found in HapMap.

Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high average quality of mapped bases.
They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
They also reported the SNPs reading significance below $ p < 10^{-6}$. 
Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against data quality \citep{Mandage2017}.

@@

\subsection{Recommendations}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
In this case, we used a cut-off of an average quality of mapped bases over 30.
Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test. (See supplementary material for lists of SNPs)

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure.

@@

\subsection{Genome wide simple logistic regression analysis}
We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}. The model we fitted to each subset of data can be expressed as such.
\begin{align*}
{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
\\
\textit{for}\quad i &= 1,\hdots, n_{p'}
\end{align*}
Here $Y_{spi}$ is a Bernoulli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$. 
We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the \todo{total number of genomic loci} , and $P_s$ is the number of populations with a segregating variant for locus $s$. In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
\caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.  }
\label{PC1_Correlation}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to quality.}
\label{MostSig}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
\caption{Logistic regression of rs6057648 for the populations with significant association to quality.}
\label{TwinsSNP}
\end{figure}

>>>>

The results section is organized as follows.
First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
Finally, we investigate how these SNPs have impacted modern genomics analyses. 
We report that major imputation servers impute the dubious SNPs. 
We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/Figure1.jpg}
\caption{
\textbf{A} 
Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. These three enrichments can be summarized as G**${\rightarrow}$GG*. (note: the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*)
\textbf{B} 
Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
The histogram on the left of the plot is the distribution of significant variants. 
\textbf{C} 
Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1kGP. This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
The same analysis was performed independently for each of the populations in the 1kGP. }
 \label{SFS}
\end{figure}

			\section{Results}

@@

When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
Regressions against different individual-level quality metrics provided by the 1kGP revealed that $Q$ per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
Individuals with low $Q$ show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that correlate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
In this case, we are using simple logistic regression looking for loci where the genotypes are predictable based on the data quality scores.
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques. 
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses, the removal of individuals with low $Q$ appears preferable to filtering positions.

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
However, the $Q$ per individual appeared to be the best metric to identify spurious sites.

@@

\caption{\textbf{A}The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project. Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once. The x-axis is ranked by the mean sequencing date per population. \textbf{B} The x-axis is sorted by the sequencing date per individual. The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. The bottom plot indicates the sequencing technologies used over time.}
\label{MapQual}
\end{figure}

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.

@@

\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
The line plot underneath shows the number of populations for which a variant has reached significance.
The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
The same variants identified as being low quality independently in each population are found in other populations. }
  \label{OverLap}
\end{figure*}

\subsection{Suspicious SNPs impact modern genomics analyses}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
We found that all of the SNPs we identified as being associated with $Q$ for chromosomes 1 and 2 were imputed. 
This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants have not been removed.

Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait. 
The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.

@@

\section{Discussion}
One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ \citep{Ebejer2013}.
While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.

Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high $Q$.
They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
They also reported the SNPs reading significance below $ p < 10^{-6}$. 
Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$ \citep{Mandage2017}.

@@

\subsection{Recommendations}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
In this case, we used a cut-off of $Q$ over 30.
Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test. (See supplementary material for lists of SNPs)

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the $Q$ scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure.

@@

\subsection{Genome wide association study}
We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}. The model we fitted to each subset of data can be expressed as such.
\begin{align*}
{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
\\
\textit{for}\quad i &= 1,\hdots, n_{p'}
\end{align*}
Here $Y_{spi}$ is a Bernoulli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$. 
We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the \todo{total number of genomic loci} , and $P_s$ is the number of populations with a segregating variant for locus $s$. In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
\caption{Regressions against the average quality per mapped bases $Q$ per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.  }
\label{PC1_Correlation}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{MostSig}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{TwinsSNP}
\end{figure}
